---
title: 'Data Science in R Final Project: Non-Residential Fire Causes'
author: "Lubov McKone"
date: "May 7, 2017"
output: html_document
---

```{r, include=FALSE}
#REQUIRED LIBRARIES
require(magrittr)
require(foreign)
require(dplyr)
require(dplyr)
require(data.table)
require(ggplot2)
require(dplyr)
require(sqldf)
require(ggplot2)
require(maptools)
require(rgeos)
require(ggmap)
require(scales)
require(RColorBrewer)
require(plotly)
```

#Introduction

The grad student we are working with, Diane, has noted that there are many fires in the NFIRS databases with "undetermined" cause, particularly after 2011.

My analysis aims to investigate how the number of non-residential fires with "undetermined" causes has changed over time and by state, and determine which years and states have statistically "out of the ordinary" number of fires with undetermined causes. 

#Data Cleaning and Subsetting

I used the fireincident data tables for every year from Diane's Google Drive. If you download these files into the working directory and name them "[year]incidents" this code will run. Unfortunately I was unable to upload them myself beacuse of their size and my time constraint.

I subsetted the fireincident data tables to include only non-residential fires, and only retained the state and "cause of ignition"" columns. I created a field with the year of the data and then used sqldf to create a table for each year listing each state and a count of fire incidents with the cause of ignition listed as "U," or undetermined. I then row-bound the tables from all of the years to create one master table listing the state, year, and number of fires with undetermined causes.

This master table is versatile.

```{r, include=FALSE}
#READ TABLES AND FILTER TO NON-RESIDENTIAL FIRES
incidents2006 <- read.dbf("2006fireincident.dbf")
incidents2006 %<>% filter(NOT_RES == "Y")
#Create a field with the year
incidents2006$INC_YEAR <- 2006
#Retain only relevant columns
incidents2006 %<>% select(STATE, FDID, CAUSE_IGN, INC_YEAR)
#Create a table that lists the state, year, and number of "unidentified" fire cause 
Ubystate2006 <- sqldf("select STATE, INC_YEAR, CAUSE_IGN, count(CAUSE_IGN) from incidents2006 where CAUSE_IGN == 'U' group by STATE, CAUSE_IGN ORDER BY STATE")

incidents2007 <- read.dbf("2007fireincident.dbf")
incidents2007 %<>% filter(NOT_RES == "Y")
#Create a field with the year
incidents2007$INC_YEAR <- 2007
#Retain only relevant columns
incidents2007 %<>% select(STATE, FDID, CAUSE_IGN, INC_YEAR)
#Create a table that lists the state, year, and number of "unidentified" fire cause 
Ubystate2007 <- sqldf("select STATE, INC_YEAR, CAUSE_IGN, count(CAUSE_IGN) from incidents2007 where CAUSE_IGN == 'U' group by STATE, CAUSE_IGN ORDER BY STATE")

incidents2008 <- read.dbf("2008fireincident.dbf")
incidents2008 %<>% filter(NOT_RES == "Y")
#Create a field with the year
incidents2008$INC_YEAR <- 2008
#Retain only relevant columns
incidents2008 %<>% select(STATE, FDID, CAUSE_IGN, INC_YEAR)
#Create a table that lists the state, year, and number of "unidentified" fire cause 
Ubystate2008 <- sqldf("select STATE, INC_YEAR, CAUSE_IGN, count(CAUSE_IGN) from incidents2008 where CAUSE_IGN == 'U' group by STATE, CAUSE_IGN ORDER BY STATE")

incidents2009 <- read.dbf("2009fireincident.dbf")
incidents2009 %<>% filter(NOT_RES == "Y")
#Create a field with the year
incidents2009$INC_YEAR <- 2009
#Retain only relevant columns
incidents2009 %<>% select(STATE, FDID, CAUSE_IGN, INC_YEAR)
#Create a table that lists the state, year, and number of "unidentified" fire cause 
Ubystate2009 <- sqldf("select STATE, INC_YEAR, CAUSE_IGN, count(CAUSE_IGN) from incidents2009 where CAUSE_IGN == 'U' group by STATE, CAUSE_IGN ORDER BY STATE")


incidents2010 <- read.dbf("2010fireincident.dbf")
incidents2010 %<>% filter(NOT_RES == "Y")
#Create a field with the year
incidents2010$INC_YEAR <- 2010
#Retain only relevant columns
incidents2010 %<>% select(STATE, FDID, CAUSE_IGN, INC_YEAR)
#Create a table that lists the state, year, and number of "unidentified" fire cause 
Ubystate2010 <- sqldf("select STATE, INC_YEAR, CAUSE_IGN, count(CAUSE_IGN) from incidents2010 where CAUSE_IGN == 'U' group by STATE, CAUSE_IGN ORDER BY STATE")

incidents2011 <- read.dbf("2011fireincident.dbf")
incidents2011 %<>% filter(NOT_RES == "Y")
#Create a field with the year
incidents2011$INC_YEAR <- 2011
#Retain only relevant columns
incidents2011 %<>% select(STATE, FDID, CAUSE_IGN, INC_YEAR)
#Create a table that lists the state, year, and number of "unidentified" fire cause 
Ubystate2011 <- sqldf("select STATE, INC_YEAR, CAUSE_IGN, count(CAUSE_IGN) from incidents2011 where CAUSE_IGN == 'U' group by STATE, CAUSE_IGN ORDER BY STATE")

incidents2012 <- read.table("2012fireincident.txt", sep="^", header=T)
incidents2012 %<>% filter(NOT_RES == "Y")
#Create a field with the year
incidents2012$INC_YEAR <- 2012
#Retain only relevant columns
incidents2012 %<>% select(STATE, FDID, CAUSE_IGN, INC_YEAR)
#Create a table that lists the state, year, and number of "unidentified" fire cause 
Ubystate2012 <- sqldf("select STATE, INC_YEAR, CAUSE_IGN, count(CAUSE_IGN) from incidents2012 where CAUSE_IGN == 'U' group by STATE, CAUSE_IGN ORDER BY STATE")


incidents2013 <- read.table("2013fireincident.txt", sep="^", header=T)
incidents2013 %<>% filter(NOT_RES == "Y")
#Create a field with the year
incidents2013$INC_YEAR <- 2013
#Retain only relevant columns
incidents2013 %<>% select(STATE, FDID, CAUSE_IGN, INC_YEAR)
#Create a table that lists the state, year, and number of "unidentified" fire cause 
Ubystate2013 <- sqldf("select STATE, INC_YEAR, CAUSE_IGN, count(CAUSE_IGN) from incidents2013 where CAUSE_IGN == 'U' group by STATE, CAUSE_IGN ORDER BY STATE")


incidents2014 <- read.table("2014fireincident.txt", sep="^", header=T)
incidents2014 %<>% filter(NOT_RES == "Y")
#Create a field with the year
incidents2014$INC_YEAR <- 2014
#Retain only relevant columns
incidents2014 %<>% select(STATE, FDID, CAUSE_IGN, INC_YEAR)
#Create a table that lists the state, year, and number of "unidentified" fire cause 
Ubystate2014 <- sqldf("select STATE, INC_YEAR, CAUSE_IGN, count(CAUSE_IGN) from incidents2014 where CAUSE_IGN == 'U' group by STATE, CAUSE_IGN ORDER BY STATE")


incidents2015 <- read.table("2015fireincident.txt", sep="^", header=T)
incidents2015 %<>% filter(NOT_RES == "Y")
#Create a field with the year
incidents2015$INC_YEAR <- 2015
#Retain only relevant columns
incidents2015 %<>% select(STATE, FDID, CAUSE_IGN, INC_YEAR)
#Create a table that lists the state, year, and number of "unidentified" fire cause 
Ubystate2015 <- sqldf("select STATE, INC_YEAR, CAUSE_IGN, count(CAUSE_IGN) from incidents2015 where CAUSE_IGN == 'U' group by STATE, CAUSE_IGN ORDER BY STATE")

#transform into data.frames
Ubystate2006 %<>% data.frame()
Ubystate2007 %<>% data.frame()
Ubystate2008 %<>% data.frame()
Ubystate2009 %<>% data.frame()
Ubystate2010 %<>% data.frame()
Ubystate2011 %<>% data.frame()
Ubystate2012 %<>% data.frame()
Ubystate2013 %<>% data.frame()
Ubystate2014 %<>% data.frame()
Ubystate2015 %<>% data.frame()

#create master table of number of undetermined causes by state and by year. 
undetermined <- rbindlist(list(Ubystate2006, Ubystate2007, Ubystate2008, Ubystate2009, Ubystate2010, Ubystate2011, Ubystate2012, Ubystate2013, Ubystate2014, Ubystate2015))

```

The table can be ordered to view the data points as time series for each state, as shown below.

| STATE | YEAR | NUM_U |
|-------|------|-------|
| AK    | 2006 | 167   |
| AK    | 2007 | 188   |
| AK    | 2008 | 153   |
| AK    | 2009 | 240   |
| AK    | 2010 | 162   |
| AK    | 2011 | 201   |
| AK    | 2012 | 201   |
| AK    | 2013 | 266   |
| AK    | 2014 | 227   |
| AK    | 2015 | 203   |

The table can also be arranged to view comparisons between states by year, as shown below:

| STATE | YEAR | NUM_U |
|-------|------|-------|
| AK    | 2006 | 167   |
| AL    | 2006 | 1081  |
| AR    | 2006 | 2172  |
| AZ    | 2006 | 855   |
| CA    | 2006 | 4651  |
| CO    | 2006 | 866   |
| CT    | 2006 | 1284  |
| DE    | 2006 | 319   |
| FL    | 2006 | 7516  |
| GA    | 2006 | 2749  |

#Analysis
To analyze the data, I create a summary table containing the mean and variance for the number of undetermined fires in each state. I decided to examine the yearly data points for Texas to see if any values were outside 2 standard deviations from the mean.

| STATE | MEAN   | STD.DEV |
|-------|--------|---------|
| TX    | 7932.6 | 1587.76 |


```{r, echo=FALSE}
tx <- read.csv("txtable.csv")
ggplot(tx, aes(YEAR, NUM_U)) + geom_ribbon(aes(ymin=7932.6-(2*1587.76), ymax=7932.6+(2*1587.76)), fill="grey87") + geom_ribbon(aes(ymin=7932.6-(1587.76), ymax=7932.6+(1587.76)), fill="grey70") + geom_line() + geom_hline(aes(yintercept=7932.6), color="coral2", size=1, linetype="dashed") + labs(x="Year", y="Number Undetermined", title="Number of Undetermined Fire Causes in Texas by Year")

```

It appears that the number of fires with undetermined causes in Texas in 2011 lies outside of 2 standard deviations from the mean (represented by the dashed coral line). Since there is only a 95% chance of this happening, something must have been going on in Texas in 2011. Further analysis could focus specifially on explaining why this number is so high, perhaps using a linear regression to understand factors that contrubte to a high number of fires with undetermined causes. 


#Shiny Web Application

To explore the data for yourself, visit this web application!

https://lmckone.shinyapps.io/FireApp/


